---
title: "NCAA 2023 WVB Analysis"
author: "Joshua Bastin"
date: "2024-1-23"
output: pdf_document
---

# Analyzing the most 'Clutch' players of the 2023 NCAA Women's Volleyball Tournament
```{r}
# Load necessary libraries for data manipulation, visualization, and table formatting
library(remotes)     # For remote package installation
library(tidyverse)   # Includes dplyr, ggplot2, and other tidyverse tools
library(datavolley)  # For handling volleyball data
library(volleyR)     # Additional volleyball analysis tools
library(knitr)       # For dynamic report generation
library(kableExtra)  # For enhanced table styling and rendering
```

```{r}
# Load DataVolley (.dvw) files and combine play-by-play data into one dataframe
dvw_files <- list.files(
  path = "/Users/joshuabastin/Dropbox/Mac/Downloads/Personal Projects/Vball Project/data",  
  pattern = "*.dvw",           
  full.names = TRUE            
)

dvw_data <- map(dvw_files, ~read_dv(.x))  # Read all .dvw files into a list
all_plays <- map(seq_along(dvw_data), ~pluck(dvw_data, .x, "plays")) %>% list_rbind()  # Extract and bind plays data
```

## Initializing team logos for visualization purposes
```{r}
# Create a dataframe of team names and corresponding logo URLs
# This will be used to display team logos in the final tables
team_logos <- data.frame(
  team = c("University of Oregon", "University of Tennessee", 
           "Creighton University", "University of Texas at Austin","University of Wisconsin-Madison",
           "University of Nebraska-Lincoln","Georgia Institute Of Technology","Stanford University","Pennsylvania State University","University of Kansas","University of Pittsburgh","University of Arkansas","Western Michigan University","Pepperdine University","University of California, Santa Barbara","University of Louisville","University of Minnesota","Arizona State University"
           ),  
  logo_url = c("https://fanapeel.com/wp-content/uploads/logo_-university-of-oregon-ducks-yellow-o-green-outline.png", "https://1000logos.net/wp-content/uploads/2021/07/Tennessee-Volunteers-logo.png", "https://upload.wikimedia.org/wikipedia/en/thumb/6/6f/Creighton_Bluejays_logo.svg/1200px-Creighton_Bluejays_logo.svg.png","https://texassports.com/images/logos/site/site.png","https://upload.wikimedia.org/wikipedia/commons/thumb/e/e5/Wisconsin_Badgers_logo.svg/2171px-Wisconsin_Badgers_logo.svg.png","https://1000logos.net/wp-content/uploads/2021/07/Nebraska-Cornhuskers-logo.png","https://upload.wikimedia.org/wikipedia/commons/thumb/b/bf/Georgia_Tech_Yellow_Jackets_logo.svg/1024px-Georgia_Tech_Yellow_Jackets_logo.svg.png","https://banner2.cleanpng.com/20180607/att/kisspng-stanford-cardinal-football-stanford-university-nca-match-day-5b19102c94b100.2516388815283691966091.jpg","https://upload.wikimedia.org/wikipedia/en/thumb/3/3a/Penn_State_Nittany_Lions_logo.svg/1200px-Penn_State_Nittany_Lions_logo.svg.png","https://upload.wikimedia.org/wikipedia/commons/9/90/Kansas_Jayhawks_1946_logo.svg","https://upload.wikimedia.org/wikipedia/commons/thumb/4/44/Pitt_Panthers_wordmark.svg/1200px-Pitt_Panthers_wordmark.svg.png","https://upload.wikimedia.org/wikipedia/commons/thumb/8/81/Arkansas_Razorbacks_logo.svg/2560px-Arkansas_Razorbacks_logo.svg.png","https://cdn.freebiesupply.com/logos/large/2x/wmu-broncos-logo-png-transparent.png","https://upload.wikimedia.org/wikipedia/commons/thumb/f/f2/Pepperdine_Waves_logo.svg/1200px-Pepperdine_Waves_logo.svg.png","https://upload.wikimedia.org/wikipedia/en/thumb/a/a8/UC_Santa_Barbara_Gauchos_logo.svg/1200px-UC_Santa_Barbara_Gauchos_logo.svg.png","https://upload.wikimedia.org/wikipedia/en/thumb/5/59/Louisville_Cardinals_logo.svg/1200px-Louisville_Cardinals_logo.svg.png","https://upload.wikimedia.org/wikipedia/commons/thumb/f/f9/Minnesota_Golden_Gophers_logo.svg/2560px-Minnesota_Golden_Gophers_logo.svg.png","https://1000logos.net/wp-content/uploads/2021/06/Arizona-State-Sun-Devils-logo.png")
)
```

## Filtering for Most Clutch Attackers
```{r}
# Filter for "clutch time" attacks (set 5 or late close-score scenarios)
# Group by player and team to calculate performance metrics:
# - kill percentage (killpct), error percentage (errpct), and hitting efficiency (hiteff)
# Include only players with at least 15 attack attempts
attacker_stats <- all_plays %>%
  filter(set_number == 5 | 
         (abs(home_score_start_of_point - visiting_score_start_of_point) <= 2 &
          home_score_start_of_point >= 17 & visiting_score_start_of_point >= 17)) %>%
  filter(skill == "Attack") %>%
  group_by(player_name, team) %>%
  summarise(
    attempts = n(),  
    kill_pct = sum(evaluation_code == "#") / n(),  
    error_pct = sum(evaluation_code %in% c("=", "/")) / n(),  
    .groups = 'drop'
  ) %>%
  mutate(hitting_eff = kill_pct - error_pct) %>%
  filter(attempts >= 15) %>%
  arrange(desc(hitting_eff)) %>%
  left_join(team_logos, by = "team")

# Select top 10 players and add ranks
top_attackers_table <- attacker_stats[1:10,]
top_attackers_table$Rank <- c(1, 2, 3, 4, 5, 6, 7, 8, 8, 10)
```

## Filtering for Most Clutch Blockers
```{r}
# Filter for "clutch time" blocks and calculate total blocks for each player
blocker_stats <- all_plays %>%
  filter(set_number == 5 | 
         (abs(home_score_start_of_point - visiting_score_start_of_point) <= 2 &
          home_score_start_of_point >= 17 & visiting_score_start_of_point >= 17)) %>%
  filter(skill == "Block") %>%
  group_by(player_name, team) %>%
  summarise(total_blocks = sum(evaluation_code == "#"), .groups = 'drop') %>%
  arrange(desc(total_blocks)) %>%
  left_join(team_logos, by = "team")

# Select top 11 blockers and add ranks
top_blockers_table <- blocker_stats[1:11,]
top_blockers_table$Rank <- c(1, 2, 3, 3, 3, 3, 3, 3, 3, 3, 3)
```

## Filtering for Most Clutch Servers
```{r}
# Filter for "clutch time" serves and calculate total aces for each player
server_stats <- all_plays %>%
  filter(set_number == 5 | 
         (abs(home_score_start_of_point - visiting_score_start_of_point) <= 2 &
          home_score_start_of_point >= 17 & visiting_score_start_of_point >= 17)) %>%
  filter(skill == "Serve") %>%
  group_by(player_name, team) %>%
  summarise(total_aces = sum(evaluation_code == "#"), .groups = 'drop') %>%
  arrange(desc(total_aces)) %>%
  left_join(team_logos, by = "team")

# Select top 8 servers and add ranks
top_servers_table <- server_stats[1:8,]
top_servers_table$Rank <- c(1, 2, 2, 2, 2, 2, 2, 2)
```

## Interesting Finding Stat
```{r}
# Find the player with the lowest hitting efficiency in "clutch time"
least_efficient_attacker <- attacker_stats %>%
  arrange(desc(hitting_eff)) %>%
  tail(1)
 # Select the last row (least efficient attacker)
```


## Creating the table for Hitting Efficiency Data
```{r}
# Create a styled table for the top clutch attackers
top_attackers_table %>%
  mutate(
    Logo = ifelse(!is.na(logo_url), 
                  paste0("<img src='", logo_url, "' style='height:20px; vertical-align:middle;'/>"), 
                  "")
  ) %>%
  select(Rank, player_name, team, Logo, attempts, kill_pct, error_pct, hitting_eff) %>%
  kbl(
    caption = "<span style='color: #333333; font-weight: bold; font-size: 20px;'>Most Clutch Attackers</span>
               <br><span style='font-size: 16px; color: #555555;'>2023 NCAA Women's Volleyball Tournament</span>
               <br><span style='font-size: 12px; color: #777777;'>Minimum 15 Attempts</span>", 
    col.names = c("Rank", "Player Name", "School", "", "Attempts", "Kill Pct", "Error Pct", "Hit Eff"),  
    align = c("c", "l", "l", "c", "c", "c", "c", "c"),  
    escape = FALSE,
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 12
  ) %>%
  column_spec(4, width = "50px") %>%  
  column_spec(1:8, color = "black") %>% 
  row_spec(0, bold = TRUE, background = "#f0f0f0", color = "black") %>%  
  footnote(
    general = "<span style='color: black;'>Data source: Volleymetrics</span>", 
    general_title = "",
    escape = FALSE 
  )
```


## Interesting Finding Table
```{r}
# Create a table for the least efficient attacker
least_efficient_attacker %>%
  mutate(
    Logo = ifelse(!is.na(logo_url), 
                  paste0("<img src='", logo_url, "' style='height:20px; vertical-align:middle;'/>"), 
                  "")
  ) %>%
  mutate(Rank = 32) %>%
  select(Rank, player_name, team, Logo, attempts, kill_pct, error_pct, hitting_eff) %>%
  kbl(
    caption = "<span style='color: #333333; font-weight: bold; font-size: 20px;'>Least Efficient Attacker in 'Clutch Time'</span>
               <br><span style='font-size: 16px; color: #555555;'>2023 NCAA Women's Volleyball Tournament</span>
               <br><span style='font-size: 12px; color: #777777;'>Minimum 15 Attempts</span>", 
    col.names = c("Rank", "Player Name", "School", "", "Attempts", "Kill Pct", "Error Pct", "Hit Eff"),
    align = c("c", "l", "l", "c", "c", "c", "c", "c"),  
    escape = FALSE, 
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 12
  ) %>%
  column_spec(4, width = "50px") %>%  
  column_spec(1:8, color = "black") %>%  
  row_spec(0, bold = TRUE, background = "#f0f0f0", color = "black") %>%  
  footnote(
    general = "<span style='color: black;'>Data source: Volleymetrics</span>", 
    general_title = "",
    escape = FALSE 
  )
```



## Creating the Blockers Table
```{r}
# Create a styled table for the top clutch blockers
top_blockers_table %>%
  mutate(
    Logo = ifelse(!is.na(logo_url), 
                  paste0("<img src='", logo_url, "' style='height:20px; vertical-align:middle;'/>"), 
                  "")
  ) %>%
  select(Rank, player_name, team, Logo, total_blocks) %>%
  kbl(
    caption = "<span style='color: #333333; font-weight: bold; font-size: 20px;'>Most Clutch Blockers</span>
               <br><span style='font-size: 16px; color: #555555;'>2023 NCAA Women's Volleyball Tournament</span>
               <br><span style='font-size: 12px; color: #777777;'>Based on Total Blocks in 'Clutch Time'</span>",
    col.names = c("Rank", "Player Name", "School", "", "Blocks"),
    align = c("c", "l", "l", "c", "c"),  
    escape = FALSE,  
    digits = 0
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 12
  ) %>%
  column_spec(4, width = "50px") %>%  
  column_spec(1:5, color = "black") %>%  
  row_spec(0, bold = TRUE, background = "#f0f0f0", color = "black") %>%  
  footnote(
    general = "<span style='color: black;'>Data source: Volleymetrics</span>", 
    general_title = "",
    escape = FALSE 
  )
```


## Creating the Serve Table
```{r}
# Create a styled table for the top clutch servers
top_servers_table %>%
  mutate(
    Logo = ifelse(!is.na(logo_url), 
                  paste0("<img src='", logo_url, "' style='height:20px; vertical-align:middle;'/>"), 
                  "")
  ) %>%
  select(Rank, player_name, team, Logo, total_aces) %>%
  kbl(
    caption = "<span style='color: #333333; font-weight: bold; font-size: 20px;'>Most Clutch Servers</span>
               <br><span style='font-size: 16px; color: #555555;'>2023 NCAA Women's Volleyball Tournament</span>
               <br><span style='font-size: 12px; color: #777777;'>Based on Total Aces in 'Clutch Time'</span>",
    col.names = c("Rank", "Player Name", "School", "", "Aces"),
    align = c("c", "l", "l", "c", "c"),  
    escape = FALSE, 
    digits = 0
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    font_size = 12
  ) %>%
  column_spec(4, width = "50px") %>%  
  column_spec(1:5, color = "black") %>%  
  row_spec(0, bold = TRUE, background = "#f0f0f0", color = "black") %>%  
  footnote(
    general = "<span style='color: black;'>Data source: Volleymetrics</span>", 
    general_title = "",
    escape = FALSE 
  )
```




